A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-05 doc/popl2013/langdef.tex
checkpoint

The following helper functions are used in some of the rules that deal with memory effects:
\begin{itemize}
\item $apply(S, E)$ is used to determine the effect on the store of a sequence of memory operations
\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
constrain how the operations in that trace can be interleaved with other sibling tasks' operations
\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
interleave the memory traces of two or more subexpressions
\end{itemize}
Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
the {\em valid\_interleave} predicate, we state some of the necessary conditions here, but leave 
the construction of actual valid interleavings until a later section.  Informally, a interleaving
of two or more memory traces must include all the operations from each trace (and no additional
operations), with each trace's operations remaining in order.  A valid interleaving must further
satisfy the requirement that the value read from the store must yield the most recently 
stored/reduced value unless that location is part of the clobber set $C$.
\end{itemize}

@@

In the case of ``read'' operations whose locations are in the clobber set $C$, we allow the local
evaluation to guess any value as long as it is of the right type.  The consistency of the value
is eventually checked by the $valid\_interleave$ predicate of some enclosing expression, but it
may not be the closest enclosing one.  The $valid\_interleave$ test itself allows some (apparent)
freedom of choice, deferring some of the consistency checks to the corresponding check further up
the expression tree.  The final case of apparent nondeterminism is in the choice of what 
additional locations to include in each subregion that results from a {\em partition} operation.
The expansion of the coloring set is necessary for the success of subsequent memory allocation
within a subregion, but that allocation might not occur within the {\em partition} expression's
body.  Although the correctness of these partitioning guesses cannot be fully verified until the
entire execution is known, when an inconsistency is encountered (i.e. there isn't a location of
the right type in the right region for a ``new $T@r$'' expression) it is possible to ``patch'' the 
proof (by adding a previously unnamed location to the appropriate expanded colorings and giving
it the right type in $H$) rather than having to start the proof over from the beginning.

@@

\begin{figure*}
\centering{
$
\begin{array}{lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \text{ where } S' = apply(S, E) \\
mark\_coherence([], \hat Q) & = & [] \\
mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
& & op(l, c', v), \text{ where } \\
& & c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases} \\
any\_interleave(E', [], \ldots, []) & = & true \\
any\_interleave([ op(l, c, v) ] \dplus E', E_1, \ldots, [ op(l, c, v) ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
valid\_interleave(S, C, E' \dplus [ read(l, c, v) ] \dplus E'', \oton{E}) & \Rightarrow & l \in C \vee S'(l) = v, \text{ where } S' = apply(S, E')
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\end{figure*}

@@

\framebox{$\opsenvx \text{newcolor }r \mapsto \emptyset, []$}

@@

K' = K \cup \{ (l_1, v_1), \ldots, (l_p, v_p) \}, \text{ where } \forall i \in [1,p]. (\forall v'. (l_i, v') \not\in K) \wedge (\forall j \in [1,p]. l_i \not= l_j) \wedge v_i \in [1,k] \\
\rho_i = \{ l : (l, i) \in K' \}, \text{ for } 1 \leq i \leq k \\

@@

\subsection{Noninterference}

@@

coherent\_interleave(S_1, S_2, C, c, [], [], []) = & true \\
coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', [ op(l, c, v) ] \dplus E_1, E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_1(l) = v) \wedge \\
& coherent\_interleave(apply(S_1, op(l, c', v)), S_2, C, c, E', E_1, E_2) \\
coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', E_1, [ op(l, c, v) ] \dplus E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_2(l) = v) \wedge \\
& coherent\_interleave(S_1, apply(S_2, op(l, c', v)), C, c, E', E_1, E_2)
\end{array}
$
\caption{work in progress}
\end{figure*}

>>>>

The following helper functions are used in some of the rules that deal with memory effects:
\begin{itemize}
\item $apply(S, E)$ is used to determine the effect on the store of a sequence of memory operations
\item $mark\_coherence(E, \hat Q)$ marks the memory operations in a subtask's trace to
constrain how the operations in that trace can be interleaved with other sibling tasks' operations
\item $valid\_interleave(S, C, E', \oton{E})$ is used to determine if $E'$ is a valid way to
interleave the memory traces of two or more subexpressions
\end{itemize}
Figure~\ref{fig:opsemfns} gives the definitions of {\em apply} and {\em mark\_coherence}.  For
the {\em valid\_interleave} predicate, we leave the precise definition to a later section.  For this
section, we need only one of the necessary conditions of a valid interleaving, that a valid
interleaving is at least an interleaving - it contains all the memory operations from each of the
constituent traces and no more, and that each trace's operations remain in order in the interleaved
trace.
\end{itemize}

@@

In the case of ``read'' operations whose locations are in the clobber set $C$, we allow the local
evaluation to guess any value as long as it is of the right type.  The consistency of the value
is eventually checked by the $valid\_interleave$ predicate of some enclosing expression, but it
may not be the closest enclosing one.  The $valid\_interleave$ test itself allows some (apparent)
freedom of choice, deferring some of the consistency checks to the corresponding check further up
the expression tree.  The final case of apparent nondeterminism is in the choice of what initial
locations to include in the result of a ``newcolor $r$'' expression.  The inclusion of locations
beyond those which are explicitly requested by the application is necessary for the success of
subsequent memory allocation within a subregion created from a partitioning using the coloring, but
that allocation might not occur within the {\em partition} expression's
body.  Although the correctness of these partitioning guesses cannot be fully verified until the
entire execution is known, when an inconsistency is encountered (i.e. there isn't a location of
the right type in the right region for a ``new $T@r$'' expression) it is possible to ``patch'' the 
proof (by adding a previously unnamed location to the appropriate expanded colorings and giving
it the right type in $H$) rather than having to start the proof over from the beginning.

@@

\begin{figure*}
\centering{
$
\begin{array}{lll}
apply(S, []) & = & S \\
apply(S, E \dplus [ read(l, c, v) ]) & = & apply(S, E) \\
apply(S, E \dplus [ write(l, c, v) ]) & = & apply(S, E)[v/l] \\
apply(S, E \dplus [ reduce_{id}(l, c, v) ]) & = & S'[id(S'(l), v)/l], \text{ where } S' = apply(S, E) \\
mark\_coherence([], \hat Q) & = & [] \\
mark\_coherence(E \dplus [ op(l, c, v) ], \hat Q) & = & mark\_coherence(E, \hat Q) \dplus \\
& & op(l, c', v), \text{ where } \\
& & c' =
\begin{cases}
simult, & \text{if }\exists \rho. l \in \rho \wedge simult(\rho) \in \hat Q \\
atomic, & \text{if }\exists \rho. l \in \rho \wedge atomic(\rho) \in \hat Q \\
excl, & \text{otherwise}
\end{cases} \\
any\_interleave([], [], \ldots, []) & = & true \\
any\_interleave([ \epsilon ] \dplus E', E_1, \ldots, [ \epsilon ] \dplus E_i, \ldots, E_n) & = & any\_interleave(E', E_1, \ldots, E_i, \ldots, E_n) \\
valid\_interleave(S, C, E', \oton{E}) & \Rightarrow & any\_interleave(E', \oton{E}) \\
\end{array}
$}
\caption{Operational Semantics Helper Functions}
\label{fig:opsemfns}
\end{figure*}

@@

\begin{array}{l@{}l}
K = \{ (l_1, iv_1), & \ldots, (l_p, v_p) \}, \text{ where } \\
& (\forall i \in [1,p]. l_i \in M(r)) \wedge \\
& (\forall i,j \in [1,p]. l_i \not= l_j)
\end{array}
{\opsenvx \text{newcolor }r \mapsto K, []}

@@

\rho_i = \{ l : (l, i) \in K \}, \text{ for } 1 \leq i \leq k \\

@@

\subsection{Coherence}

The soundness proof above guarantees that the evaluation of an expression produces a value of the
right type, and do so within the regions of memory for which privileges are held, but for the
Legion programming model to be useful, we need to also guarantee that the evaluation of an expression
produces the (or at least, a) correct value.  The key to doing this is to place additional 
constraints on what it means to be a valid interleaving of two or more memory traces.  In
addition to a valid interleaving being an interleaving of the individual traces, we will 
require that it be:
\begin{itemize}
\item {\em coherent} - each memory read operation should see the effect of all writes and reductions 
that occur earlier in the interleaved trace to the same location, and
\item {\em sequentially equivalent} - the value read by each memory read operation and the result of applying the interleaved trace to the original store should be the same as if each constituent trace
was applied to the store in a valid {\em program order}.
\end{itemize}
These requirements are relaxed in some cases due to either a non-empty clobber set (representing
operations that may occur concurrently as part of evaluating other functions) or due to coherence
modes other than {\em excl} on operations with the traces (representing operations that may 
occur concurrently as part of evaluating other parts of the same function).

Figure~\ref{fig:validinterleave} gives the precise definition of the {\em valid\_interleave}
predicate, but there are a lot of moving parts, so description in ``plain'' English may help.
Consider a set of constituent traces $\oton{E}$.  Let the {\em exclusive location set} $L_{excl}$ be
the set of all locations that appear in at least one operation with coherence mode $excl$ in at least
one constituent trace, excepting any locations in the clobber set $C$.  Let the {\em atomic location
set} $L_{atomic}$ similarly be the set of locations that appear in at least one operation with 
coherence mode $atomic$, but no operations with coherence mode $excl$, again excepting any location
in the clobber set.  A memory trace $E'$ is considered a valid interleaving of $\oton{E}$ if all of
the following are true:
\begin{enumerate}
\item $E'$ is an interleaving of $\oton{E}$.
\item For every read in $E'$ with a location in $L_{excl}$, the value read is what would be in the
 store at that location if all the previous operations of $E'$ were applied to the original
store $S$.
\item For every read in $E_i$ with a location in $L_{excl}$, the value read is what would be in the
store at that location if all preceding constituent
traces (i.e. $E_1, \ldots, E_{i-1}$) were applied to the store $S$, followed by the operations from
$E_i$ that precede the read.
\item For every location $l$ in $L_{excl}$, the value
at that location after the trace $E'$ is applied to $S$ must match the value that would be in
the store if each of $E_i$ were successively applied to $S$.
\item For every read in $E'$ with a location in $L_{atomic}$ for which a write operation to the same
location appears earlier in $E'$, the value read is what would be in the store if all the previous
operations of $E'$ were applied to the initial store $S$.
\item There is some permutation $\oton{\pi}$ of the integers $1,\ldots,n$ such that, for every read
in $E_{\pi_i}$ with a location in $L_{atomic}$, the value read is what would be in the store at that
location if the traces $E_{\pi_1}, \ldots E_{\pi_{i-1}}$ were applied to $S$, following by the operations 
from $E_{\pi_i}$ that precede the read.
\item Using the same permutation $\oton{\pi}$, for every location in $L_{atomic}$, the value at that
location after the trace $E'$ is applied to $S$ must match the value that would be in the store
if each of $E_{\pi_i}$ were successively applied to $S$.
\end{enumerate}

Flexibility in interleaving operations is the key to achieving parallel execution, but
hopefully the reader will agree that the exact criteria for creating a valid interleaving is
much too expensive to implement at runtime, so we now consider whether there are simpler ways to
generate valid interleavings, and under what conditions those approaches may be used.

The first option available to us is to serialize the constituent traces into the original
program order.  Although each constituent trace can be assumed to be coherent, that coherence is
in the context of a clobber set that may differ from the one in which the interleaving must occur.
As discussed earlier, this difference can result in cases where no valid interleaving exists.
However, we can show that as long as any valid interleavings exist, the serialization of the
constituent traces will be one of them.

To prove this result, we will use the following lemma:

\begin{lem}
\label{lemma:applychain}
For any initial store $S$ and memory traces $E_1$ and $E_2$,
$$apply(S, E_1 \dplus E_2) = apply(apply(S, E_1), E_2).$$
\end{lem}

The proof of this follows directly from the definition of $apply$.

\begin{thm}
Let $S_0$ be an initial store, $C$ be a clobber set and $\oton{E}$ be memory traces to be
interleaved.  Let $E' = E_1 \dplus \ldots \dplus E_n$.  If there exists any $E''$ that is a
valid interleaving of $\oton{E}$, then $E'$ is also valid interleaving of $\oton{E}$.
\end{thm}

\begin{proof}
To prove $valid\_interleave(S_0, C, E', \oton{E})$, we need to show the three conditions hold:
\begin{enumerate}
\item $any\_interleave(E', \oton{E})$ is intuitively obvious, and can be proven by induction over 
the length of $E_1$, followed by $E_2$, and so on.
\item $coherent(S_0, C, E')$ is shown by considering each constituent trace's subsequence in $E'$
in order, using Lemma~\ref{lemma:applychain} along with the coherence of each $E_i$.
\item $seq\_equiv(S_0, C, E', \oton{E})$ is trivial, as each leading subsequence of $E'$ corresponds
to the appropriate pieces of $\otok{i}{E}$.
\item $coherent(S_0, C, E')$ can be shown from the sequential equivalence of $E''$.  The value 
returned by each read in $E''$ has been shown to be equal to the value in the store that results
from the application of all previous constituent traces and the operations before the read in its
own trace.  This is exactly the sequence of preceding operations of the corresponding read in $E'$,
and Lemma~\ref{lemma:applychain} allows us to trainsform the chain of $apply$ operations into a 
single $apply$ operation on the whole sequence, satisfying the coherence condition for the read
in $E'$.  This property applies for all reads in $E''$, and since both $E'$ and $E''$ are
interleavings of the same traces, they have exactly the same reads, so we have shown all the reads
in $E'$ to be coherent.
\end{enumerate}
\end{proof}

Now that we have a way to construct a valid interleaving, we can look at how to modify that
interleaving to get interleavings that are still valid, but allow parallel execution.  To do this,
we consider the criteria under which we may swap the order of two adjacent memory operations in an
existing valid interleaving.
\begin{itemize}
\item First, the two operations must have come from different constituent
traces, or the swap would violate the $any\_interleave$ criterion.
\item Second, if one is a read of
a location (that is not in the clobber set), the other may not be a write or reduce to the same
location, as changing the order would almost certainly change the value that is read.
\item Third, if one is a write of a location, the other may not be a write or reduce to the same
location, as the store that would result from applying those operations in the opposite order would
show a difference in that location.
\item Finally, if both operations are reductions to the same location, they may be swapped if they
use the same reduction function, as reduction functions are assumed to be commutative.  (We 
follow the example of other work (i.e. DPJ) and leave proofs of commutativity to the programmer,
possibly with the aid of other tools.)  However, if they use different reduction functions (by
name - we make no attempt to compare the operation of functions), commutative cannot be 
guaranteed and swapping the two operations might result in a different resulting store.
\end{itemize}

To summarize this, we define a {\em noninterference} operator on two effects that have been
annotated with which constituent trace they came from as follows:
$$
\begin{array}{l@{ }l}
op_1^{E_1}(l_1, c_1,& v_1)\ \#\ op_2^{E_2}(l_2, c_2, v_2) \Leftrightarrow E_1 \not= E_2\ \wedge \\
& \big( l_1 \not= l_2 \vee (op_1 = read \wedge op_2 = read)\ \vee \\
& (op_1 = reduce_{id_1} \wedge op_2 = reduce_{id_2} \wedge id_1 = id_2) \big)
\end{array}
$$

\begin{lem}
Let $S$ be an initial store, $C$ be a clobber set, $\oton{E}, E'_a, E'_b$ be memory traces, and
$\epsilon_1, \epsilon_2$ be two annotated memory operations such that $\epsilon_1 \# \epsilon_2$.
Then,
$$valid\_interleave(S, C, E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b, \oton{E})$$
if and only if
$$valid\_interleave(S, C, E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b, \oton{E}).$$
\end{lem}

\begin{proof}
As before, we consider the three conditions needed to be a valid interleaving.  We describe the
implication in the ``forward'' direction, but the arguments are all symmetric.
\begin{enumerate}
\item If $E'_a \dplus \epsilon_1 \dplus \epsilon_2 \dplus E'_b$ is an interleaving of $\oton{E}$,
the proof must first ``pop off'' all of $E'_a$, leaving $\epsilon_1 \dplus \epsilon_2 \dplus E'_b$ for
the interleaved subtrace.  The next two steps must pop $\epsilon_1$ and $\epsilon_2$, leaving $E'_b$.
They cannot be in the same subtrace because $\epsilon_1 \# \epsilon_2$, so we could also pop
$\epsilon_2$ and then $\epsilon_1$, resulting in a proof that $E'_a \dplus \epsilon_2 \dplus \epsilon_1 \dplus E'_b$ is also an interleaving of the constituent traces.
\item For coherence, we must show the value read (if any) by $\epsilon_1$ does not change, the value
read (if any) by $\epsilon_2$ does not change, and that any read in $E'_b$ sees no change.  With the
help of Lemma~\ref{lemma:applychain}, we can eliminate the common $E'_a$ from all sequences by
using $S' = apply(S, E'_a)$.  Similarly to show that $apply(S', \epsilon_1 \dplus \epsilon_2 \dplus E) = apply(S', \epsilon_2 \dplus \epsilon_1 \dplus E)$ for any $E$, we need only to show that it holds
for $E = []$.  The proof is now reduced to showing the following:
\begin{enumerate}
\item $\epsilon_1 = read(l, c, v) \Rightarrow S'(l) = apply(S',\epsilon_2)(l)$ - since $\epsilon_1$
is a read, either $\epsilon_2$ must be a read (in which case $apply(S, \epsilon_2) = S'$) or must be
to a different location, so the result of applying just it to $S'$ cannot change the value in 
location $l$.
\item $\epsilon_2 = read(l, c, v) \Rightarrow apply(S', \epsilon_1)(l) = S'(l)$ - this is the same
as above, with $\epsilon_1$ and $\epsilon_2$ switched.
\item $apply(S', \epsilon_1 \dplus \epsilon_2) = apply(S', \epsilon_2 \dplus \epsilon_1)$ - we must
show that the two store operations $apply(\bullet, \epsilon_1)$ and $apply(\bullet, \epsilon_2)$
commute.  Modifications to two different locations commute, as do two reads (which make no change
to the store).  Finally, two reductions using the same function to the same location also commute.
\end{enumerate}
\item The proof that sequential equivalence is maintained comes down to showing the final store
is identical after the swap of the two operations, and this is merely a special case of the third
piece of the coherence proof.
\end{enumerate}
\end{proof}

First, we would like a fallback approach that will always result in a valid interleaving, even if
we lack information about the constituent traces (or perhaps lack the time to analyze it).  The

@@

coherent\_interleave(S, C, c, E', [], E_1, [], E_2, []) = & \forall l. l \in C \vee apply(S, E')(l) = apply(S, E_1 \dplus E_2)(l) \\
coherent\_interleave(S, C, c, E'_a, \epsilon \dplus E'_b, E_{1a}, \epsilon \dplus E_{1b}, E_{2a}, E_{2b}) = & (\epsilon = read(l, c, v) \wedge l \not\in C \Rightarrow apply(S,E'_a)(l) = v \wedge apply(S,E_{1a})(l) = v) \wedge \\
& coherent\_interleave(S, C, c,  E'_a \dplus \epsilon, E'_b, E_{1a} \dplus \epsilon, E_{1b}, E_{2a}, E_{2b}) \\
coherent\_interleave(S_1, S_2, C, c, [ op(l, c', v) ] \dplus E', E_1, [ op(l, c, v) ] \dplus E_2) = & (op = read \wedge l \not\in C \wedge c = c' \Rightarrow S_2(l) = v) \wedge \\
& coherent\_interleave(S_1, apply(S_2, op(l, c', v)), C, c, E', E_1, E_2)
\end{array}
$
\caption{work in progress}
\end{figure*}
